Sources of Measurement Bias in Modern CPUs

### 2014-02-12 paper.tex
Cleanup and small changes

\subsection{Analysis}
To be able to explain what is happening in this case we need to know the addresses of all variables at run time.
For statically allocated data, memory allocation is decided at compile time.
The location of variables i, j and k can be found by looking at the symbol table in the ELF executable.
Below is an excerpt of \lstinline$readelf -s$.
\begin{lstlisting}[float=h, basicstyle=\small]
37: 000000000060103c    4  (...)  25 i
38: 0000000000601040    4  (...)  25 j
39: 0000000000601044    4  (...)  25 k
\end{lstlisting}

@@

\section{Bias from Heap Allocations}
Address aliasing can be caused by conflicting pairs of load/store operations to any part of memory.
The previous section looked specifically at an example of collision between stack variables and static data.
Another scenario to consider is collisions in heap allocated memory.
In particular, code that operate on pairs of contiguous arrays can be vulnerable to 4K aliasing.
Acquiring dynamic memory at run time is usually done by calling malloc, which takes a number of bytes to allocate and returns a pointer to that area.
A typical malloc implementation uses two different strategies for how to allocate memory.
The \emph{heap} area in virtual memory is used for smaller allocations.
Programs are initially given a small heap area, which the allocator can increase by calling brk or sbrk.
For larger allocations, it is typically faster to use anonymous memory mapping by calling mmap, instead of growing the ``normal'' heap.
The heap starts at a low address close to the data section and grows towards higher addresses. 
Mmap allocates much closer to the stack.
While heap addresses can look like 0x16e30a0 or 0x1723020, pointers returned by mmap can for example be 0x7f0318a8f010 or 0x7f03105d2010.
This distinction is of course unimportant for application developers, as everything is conceptually the same ``heap''.
However, mmap has an interesting property in that allocations will always be page aligned.
The page size is 4096 bytes, meaning two pointers returned by mmap will \emph{always} alias.\footnote{libc's version of malloc adds 16 bytes of medatada at the beginning, therefore every mmap'ed address ends with 0x010}

@@

\subsection{Bias from Heap Allocators}
Addresses of dynamic memory at run time is governed by the particular heap allocator used.
Like the contents of environment varialbes, the exact operation of heap allocation can typically not be known ahead of time.
For the code examples we present in C, the dynamic linker will resolve malloc to whichever version of libc is available on the system.
In other words, it is part of the environment, and a potential source of measurement bias.
(table, todo)

@@

We use the currently latest version 3.10.1 of ATLAS, built as a shared library from source.
The automatic tuning happens during the build process.
A series of test programs are run to determine cache edges and other properties of the hardware, which in turn affects the resulting binary.

>>>>

\subsection{Finding Collisions}
To be able to explain what is happening in this case we need to know the addresses of all variables at run time.
For statically allocated data, memory allocation is decided at compile time.
The location of variables i, j and k can be found by looking at the symbol table in the ELF executable.
Below is an excerpt of \lstinline$readelf -s$.
\begin{lstlisting}[float=h, basicstyle=\small]
37: 000000000060103c    4  (...)  25 i
38: 0000000000601040    4  (...)  25 j
39: 0000000000601044    4  (...)  25 k
\end{lstlisting}

@@

\section{Bias from Heap Allocators}
Address aliasing can be caused by conflicting pairs of load/store operations to any part of memory.
The previous section looked specifically at an example of collision between stack variables and static data.
Another scenario to consider is collisions in heap allocated memory.
In particular, code that operate on pairs of contiguous arrays can be vulnerable to 4K aliasing.
Acquiring dynamic memory at run time is usually done by calling malloc, which takes a number of bytes to allocate and returns a pointer to that area.
A typical malloc implementation uses two different strategies for how to allocate memory.
The \emph{heap} area in virtual memory is used for smaller allocations.
Programs are initially given a small heap area, which the allocator can increase by calling brk or sbrk.
For larger allocations, it is typically faster to use anonymous memory mapping by calling mmap, instead of growing the ``normal'' heap.
The heap starts at a low address close to the data section and grows towards higher addresses. 
Mmap allocates much closer to the stack.
While heap addresses can look like 0x16e30a0 or 0x1723020, pointers returned by mmap can for example be 0x7f0318a8f010 or 0x7f03105d2010.
This distinction is of course unimportant for application developers, as everything is conceptually the same ``heap''.
However, mmap has an interesting property in that allocations will always be page aligned.
The page size is 4096 bytes, meaning two pointers returned by mmap will \emph{always} alias.\footnote{libc's version of malloc adds 16 bytes of medatada at the beginning, therefore every mmap'ed address ends with 0x010}

@@

\subsection{Behavior of other Allocators}
Addresses of dynamic memory at run time is governed by the particular heap allocator used.
Like the contents of environment varialbes, the exact operation of heap allocation can typically not be known ahead of time.
For the code examples we present in C, the dynamic linker will resolve malloc to whichever version of libc is available on the system.
In other words, it is part of the environment, and a potential source of measurement bias.
(table, todo)

@@

We use the currently latest stable version 3.10.1 of ATLAS, built as a shared library from source.
The automatic tuning happens during the build process.
A series of test programs are run to determine cache edges and other properties of the hardware, which in turn affects the resulting binary.
